---
title: "Exploratory DNA Methylation Analysis"
author: "Abdullah Hanif"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 5)
results_dir <- "../results"
metadata_dir <- "../data/metadata"
qc <- read.delim(file.path(results_dir, "qc_summary.tsv"), check.names = FALSE)
sex <- read.delim(file.path(results_dir, "sex_check.tsv"), check.names = FALSE)
filters <- read.delim(file.path(results_dir, "probe_filter_counts.tsv"), check.names = FALSE)
diff_summary <- read.delim(file.path(results_dir, "differential_summary.tsv"), check.names = FALSE)
dmr_summary <- read.delim(file.path(results_dir, "dmr_summary.tsv"), check.names = FALSE)
enrichment_summary <- read.delim(file.path(results_dir, "enrichment_summary.tsv"), check.names = FALSE)
enrichment <- read.delim(file.path(results_dir, "enrichment_results.tsv"), check.names = FALSE)
metadata <- read.delim(file.path(metadata_dir, "sample_sheet_clean.tsv"), check.names = FALSE)
```

# Aim and data

This report demonstrates a reproducible workflow for exploratory human brain DNA methylation array analysis in an Alzheimer's disease context. It is a methods demonstration, not a claim of biological discovery.

The data are from GEO accession GSE212682 and were generated with an Illumina MethylationEPIC array. The analysis uses the selected brain region and the fixed, balanced subset described in the repository README. Raw IDAT files are not included in the repository.

```{r sample-summary}
knitr::kable(as.data.frame(table(metadata$diagnosis)),
             col.names = c("Diagnosis", "Samples"))
```

The selected metadata fields and their observed values are shown below. Age is reported as supplied by GEO; no numeric age was invented when the source value was `90+`.

```{r metadata-summary}
knitr::kable(data.frame(
  variable = c("region", "diagnosis", "sex", "batch", "bisulfite_batch", "age_raw"),
  observed_values = c(
    paste(unique(metadata$region), collapse = "; "),
    paste(unique(metadata$diagnosis), collapse = "; "),
    paste(unique(metadata$sex), collapse = "; "),
    paste(unique(metadata$batch), collapse = "; "),
    paste(unique(metadata$bisulfite_batch), collapse = "; "),
    paste(unique(metadata$age_raw), collapse = "; ")
  )
))
```

# QC and normalization

Mean detection p-values were used as a sample-level assay performance check. The threshold and observed results below come directly from the QC output.

```{r qc-summary}
knitr::kable(qc)
knitr::kable(data.frame(
  samples = nrow(qc),
  samples_failed = sum(qc$qc_status == "fail"),
  sex_mismatches = sum(sex$sex_mismatch, na.rm = TRUE)
))
```

![Sample detection p-values](../results/qc_mean_detection_p.pdf)

![Intensity QC](../results/qc_intensity.pdf)

Sex prediction was compared with the supplied metadata. A mismatch is a flag for investigation, not proof that either field is correct.

```{r sex-summary}
knitr::kable(sex)
```

Noob normalization was used to correct background signal and dye-related technical differences. The raw-versus-Noob beta density comparison was generated from the normalized objects.

![Raw versus Noob beta distributions](../results/normalisation_beta_density.pdf)

# Probe filtering and missingness

The filtering counts below are generated by code. Sex-chromosome probes and probes with annotated CpG or single-base-extension SNPs were removed. A validated EPIC cross-reactive list was not available in the allowed package set, so that filter was not silently substituted.

```{r filtering}
knitr::kable(filters)
knitr::kable(enrichment_summary)
```

The final beta matrix was checked for missing values. No imputation was performed when no missing beta values were observed. This is different from genotype imputation for SNP arrays.

# PCA and batch assessment

PCA was used to inspect the largest axes of variation. Batch was visualized but not removed blindly. In this subset, age has no variation and diagnosis is confounded with the observed batch groups; therefore the primary differential model does not include batch.

![PCA by diagnosis](../results/pca_by_diagnosis.pdf)

![PCA by batch](../results/pca_by_batch.pdf)

```{r pca-tables}
pca_variance <- read.delim(file.path(results_dir, "pca_variance.tsv"), check.names = FALSE)
pc_tests <- read.delim(file.path(results_dir, "pc_variable_associations.tsv"), check.names = FALSE)
knitr::kable(head(pca_variance, 5))
knitr::kable(pc_tests)
```

# Differential methylation

Limma testing was performed on M-values. Delta-beta values are reported for interpretation on the beta scale. The model summary and the first rows of the generated table are shown below.

```{r differential}
knitr::kable(diff_summary)
differential <- read.delim(file.path(results_dir, "differential_top.tsv"), check.names = FALSE)
knitr::kable(head(differential, 10))
```

![Differential methylation volcano plot](../results/differential_volcano.pdf)

![Differential methylation QQ plot](../results/differential_qq.pdf)

# DMRcate

DMRcate was run with EPICv1 annotation, a 1000 base-pair smoothing window, at least two consecutive CpGs, and an FDR cutoff. The result is exploratory and strongly dependent on the small sample and confounded design.

```{r dmr}
knitr::kable(dmr_summary)
dmr_ranges <- read.delim(file.path(results_dir, "dmr_ranges.tsv"), check.names = FALSE)
knitr::kable(head(dmr_ranges, 10))
```

# Enrichment

`missMethyl::gometh` was used because it accounts for the unequal number of array probes mapping to genes. The strict differential list was small, so the script used a code-derived ranked set of the top CpGs by limma p-value, with all filtered CpGs as the background. The result is exploratory and is not evidence of pathway discovery.

```{r enrichment}
knitr::kable(enrichment_summary)
knitr::kable(head(enrichment, 10))
```

![Top exploratory pathways](../results/enrichment_top_pathways.pdf)

# Limitations

- The raw EPIC analysis uses a small, laptop-memory-constrained subset.
- Diagnosis and batch are confounded in this subset.
- Age is supplied as `90+` and has no usable numeric variation here.
- Brain tissue is bulk tissue with unknown cell-type composition.
- There is no replication cohort in this demonstration.
- The differential and enrichment results are exploratory.
- A validated EPIC cross-reactive-probe list was not available in the allowed package set.
- The full differential table is kept locally because it exceeds GitHub's file-size limit.

# Session information

```{r session-info}
sessionInfo()
```
